\propertitle

### 2019-12-06 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-06T12:02:48+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Answering this kind of questions always engages an element of uncertainty.
We cannot say \enquote{the answer is such and such}; at best we can assign
degrees of reasonable belief -- that is, probabilities -- to the possible
answers. The assessment of such distributions of belief involves
experimental data, such as recordings of neuronal activity from specific
brain areas, and pre-data knowledge and assumptions about biological
conditions and mechanisms. Our pre-data \dobs\ are often simplified, to be
mathematically tractable, and are therefore called \enquote{models}.

The best experimental measurements of instantaneous neuronal activity use
remarkable technologies, but can still only record a very small sample of
neurons -- hundreds at most -- compared to the numbers that constitute a
functionally distinguished brain region. Many probabilistic models focus on
such samples only: they somehow neglect, in their mathematical assumptions,
that the recorded neurons are a sample from a larger population. Such
isolating assumptions sometimes escape attention, being subtly hidden in
the mathematics. Some probabilistic models try to take unrecorded neurons
into account -- but by describing each neuron individually, thus becoming
very complex \citep{huang2015,battistinetal2017}. It would be useful to
explore models that operate in between, addressing the larger brain area
whence the sample comes, but collectively, without asking about individual
neuronal details. Such intermediate models would be useful for preliminary
investigations, to help us to decide which hypotheses to discard and which
to consider for more complex and costly studies, or to suggest new
hypotheses.

% \mynote{[Yasser:] In the present work we aim to address the issue of
%   building probability distributions over a large population using
%   recording from subpopulations by considering the distribution of total
%   population activity as an example. [} 
In the present work we propose such
an intermediate probability model. It answers this question: How much was
the \emph{total} activity of a large neuronal population, given the
observation of the activity of a very small sample thereof? This model
addresses a larger brain area, avoiding the assumption of isolation of the
sample; and by focusing on the total activity, rather than the activity of
individual neurons, it remains simple and numerically tractable.

The model we propose is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b_r1996,grandy1980}
has been used for different kinds of investigations about the neuronal
activity of various brain areas and about other phenomena of importance to
the neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.

We illustrate possible uses of our proposed model in
\sect~\ref{sec:application}, by applying it to two concrete data sets:
\begin{enumerate*}[label=(\alph*)]
\item the activity of 65 neurons recorded for 20 min from a rat's Medial
  Entorhinal Cortex \citep{stensolaetal2012}, \item the activity of 159
  neurons recorded for 15 min from a macaque's Motor Cortex
  \citep{rostamietal2016_r2017}.
\end{enumerate*}
For each data set the model gives us the most plausible frequencies with
which all levels of total activity of a much larger population,
1\,000--10\,000 neurons, appeared during the recording. For example it can
tell us that 60 out of 10\,000 neurons were likely active during 1\% of the
recording time (though not necessarily always the same 60), 250 neurons out
of 10\,000 were active during 0.4\% of the time, and so on. The precise
meaning of this frequency distribution is explained in
\sect~\ref{sec:model}. For the two example data sets, the guessed frequency
distributions in the larger population are distinctly different from those
in the sample. For the first data set the frequency distribution for the
larger population has two very distinct modes, both at low
activities %($0.7\%$ and $2.6\%$ of simultaneously active neurons
% within $3\;\mathrm{ms}$,
(see \fig~\ref{fig:plots_5mom}), whereas the frequency distribution for the
sample is monotonically decreasing with its maximum at zero activity. The
frequency distribution for the second data set presents one prominent
shoulder in its low-activity mode. These results show that the proposed
model can lead to the formulation or preliminary assessment of interesting
hypotheses, as we will illustrate with a toy example. Note that these
guessed features of the full population could not be inferred by the
application of maximum-entropy \emph{to the sample alone}.

@@

Our model also solves a methodological problem in the use of
maximum-entropy methods to assess the \enquote{cooperativity},
\enquote{interaction}, or \enquote{synchrony} in neuronal activity, for
example studying its pairwise correlations and correlations of higher order
\citep[see for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
When the difference in size between a large population and a sample thereof
is too large, the presence of some inferential properties of correlations
for the sample implies the \emph{lack} of such properties for the larger
population, and vice versa. Maximum-entropy applications at the sample
level can therefore deceive us in questions regarding the cooperativity of
the larger population. The application to the larger population is more
reliable. We discuss this point in detail in
\sect~\ref{sec:marginalization}.

@@

\section{Quantifying the importance of higher-order correlations:\\
  the limitations of methods at the sample level}
\label{sec:marginalization}

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been used as a way of
quantifying the \enquote{cooperativity} \citep[\eg][]{gersteinetal1985} or
\enquote{interaction}
\citep[\eg][]{martignonetal1995,schneidmanetal2006,shlensetal2006} or
\enquote{synchrony} \citep[\eg][-- we're only citing early papers using
these terms]{bohteetal2000,amarietal2003} of neuronal activity. Most, if
not all, such applications apply the method \emph{at the sample level}. In
this section we discuss how our proposed application bears on this kind of
quantification and compare it with the traditional application at the
sample level. But such a comparison involves some important methodological
caveats which we wish to discuss first.% For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our
% discussion holds for any sets of statistics.

@@

This fact is important for our analysis. If, say, means and pairwise
correlations seem informationally sufficient for a particular sample from a
brain area, then they may well not be sufficient for the larger population
of neurons constituting that area, and vice versa. So if we are interested
in the \enquote{cooperativity} or \enquote{interaction} of a brain area, it
is unreliable to use a maximum-entropy distribution constructed only for a
sample
thereof. % \citep[contrast this with some literature, \eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}
The approach presented here avoids this problem because the maximum-entropy
method is applied to obtain the distribution of the larger population, not
of the sample alone.

@@

Compare the results above with those obtained by applying the method at the
sample level, that is, with $N=n$:
\begin{equation}
  \label{eq:diff_suff_n}
  \begin{split}
  \ydi_{n}(M_{4},M_{2}) &= 280\;\nat = 1\,220\;\hart,
\\    \ydi_{n}(M_{5},M_{4}) &= 3.3\;\nat = 1.4\;\hart.
\end{split}
\end{equation}
The data are 1\,220 orders of magnitude more probable conditional on four
moments than conditional on two, and $10^{1.4}=25$ times more probable
conditional on five moments than on four moments.

@@

The crucial point is our belief distribution for the activity of the
sample conditional on the activity of the larger population, given by
formula~\eqref{eq:conditional_hypergeometric} in
appendix~\ref{sec:derivation_maxent_fullnet}, reproduced here:
\begin{equation}
  \labelbis{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
    \binom{\yA}{\ya}\binom{N-\yA}{n-\ya}\binom{N}{n}^{-1}
\equiv \binom{n}{\ya}\binom{N-n}{\yA-\ya}\binom{N}{\yA}^{-1}
\end{equation}
leading to the equality of factorial
moments~\eqref{eq:equal_m_factorialmoments}, on which our approach rests.
This conditional probability, characteristic of \enquote{drawing without
  replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968},
ensues when our degree of belief that the $i$th unit (ball or neuron) will
be drawn has the same value for all $i \in \set{1,\dotsc,N}$. Thus,
consider the pool of all neurons which we believe -- with equal degree for
each neuron -- could have been recorded. $N$ is the size of that pool. $N$
therefore depends on factors such as: the shape, dimensions, and technical
specifications of the recording probe; the inaccuracy in the insertion of
the probe, leading for example to slightly different insertion points or
angles; the density of neurons around the probe. But it also depends on the
homogeneity of the brain region where the recording was made: if we believe
that it doesn't matter whether the probe had been inserted in some other
point of the same brain region, then
formula~\eqref{eq:conditional_hypergeometric} is appropriate. $N$ can
therefore only be assessed case by case.

@@

A rigorous assessment would involve a more expensive, full Bayesian
calculation; but such calculation would make the whole maximum-entropy
approach superfluous. We discuss this final point in the next section.

@@

The possible uses of the method here presented depend on the assumptions
and approximations on which it rests: three assumptions and two
approximation all in all. These become clear within a full-fledged
probabilistic approach. We summarize them here and give a more mathematical
sketch in appendix~\ref{sec:inference_prob}.

@@

The second assumption, typical of maximum-entropy in general, concerns the
form of the probability distribution for the long-run frequencies ensuing
from the first assumption. It is taken to be either an \enquote{entropic
  prior} \citep{skillingetal1984,rodriguez1991,neumann2007}, which takes
into account the multiplicity of long-run frequencies, or a prior for a
model with sufficient statistics, which expresses the sufficiency of the
moments under study \citep{portamana2017}. This assumption is crucial for
obtaining the maximum-entropy method with the Shannon entropy. A different
density leads to alternative maximum-entropy methods, for example based on
the Burg entropy \citep{burg1975} if a Dirichlet density is used
\citep{jaynes1986d_r1996}. The entropic prior depends on a reference
distribution and on a coefficient $L$ which determines the sharpness of the
prior's peak. The computations of the preceding sections were insensitive
to the reference distribution in their significant digits. We used three
biologically reasonable alternatives: a flat distribution, one linearly
decreasing with $\yA$, and a normal one with a broad peak at low
activities.

@@

Finally, the maximum-entropy method typically approximates the number $T$
of time bins, the prior coefficient $L$, and the ratio $T/L$ to infinity.
These approximations lead to the identification of the four distributions
listed at the end of \sect~\ref{sec:model}: the recorded frequency
distribution becomes equal to the long-run one, and also to the probability
distribution for the activity at recorded and new time bins. In our
application it also leads to the equality of the hypergeometric
distribution~\eqref{eq:conditional_hypergeometric} and the conditional
frequencies of the sample activity $\ya$ given $\yA$. The goodness of this
approximation decreases as the ratio between the number of possible joint
states and the number of bins -- roughly $nN/T$ -- increases, and also as
observed quantities approach their mathematical bounds, such as some
frequencies $\yff_{\ya}=0$. Without these approximations it is still
possible to interpret the maximum-entropy
distribution~\eqref{eq:app_maxent_pop_first} as the mode of the probability
distribution for the recorded frequencies $\yF$, that is, the most probable
recorded frequency.

@@

The $T= \infty$ approximation can break the method for specific
distributions of sample frequencies, if too many moments are used. For
example take $n=2$ with frequency distribution $\yf=(0.5,0,0.5)$ for the
activities $\ya=0,1,2$; that is, the sample is silent half of the time and
completely active half of the time. The first two normalized factorial
moments have both value $0.5$, and together they uniquely determine $\yf$.
If we now take $N=3$, it can be proved that
\eqn~\eqref{eq:app_maxent_pop_first}, constrained on those two moments, has
no solution. The reason is clear from a sampling perspective: if there is
zero probability of sampling one active neuron, then the larger population
cannot have any active neurons, and there must be zero probability of
sampling two active ones. But this reasoning holds only for each single
time bin, not for the averages over time bins. The $T=\infty$ approximation
identifies the two instead. Outside this approximation there are no
contradictions because the conditional frequencies and conditional
probabilities are kept distinct.

@@

The reason why we record neurons from a brain region is to have an idea of
the activity of the other neurons in that region, similarly to survey
sampling. If we thought that \emph{precisely} the recorded ones were
exceptions or completely unrelated to the others surrounding the recording
probe, our recording would serve little purpose. Probabilistic methods that
allow us to make direct inferences about the larger population are
therefore useful. But they become complex and sensitive if we ask for very
detailed inferences.

We have presented a method that allows us to make direct inferences about a
larger population from very small samples. Its inferences regard only the
frequencies of the total activity of the larger population, but such
limited scope makes the method computationally very cheap.

Yet, and most important, the method leads to inferences that are not at all
trivial and not easily discernible by looking at the sample. We showed this
by applying the method to two real data sets.

By \enquote{inference} we mean the quantification of a degree of belief
about possible observations, based on specific assumptions; not the
response of an oracle. The method gives us a \emph{forecast} based on those
assumptions, and therefore has two main uses, just like any other
probabilistic method. If our assumptions are only hypotheses under testing,
then the subsequent confirmation or confutation of the forecast will
increase or decrease our confidence in those hypotheses. If we strongly
trust our assumptions, on the other hand, then we rely on the forecast and
even use it in place of actual observations, especially if the latter are
difficult to make. (In fact, we typically shift from the first to the
second use.) In the preceding sections we have given examples of hypotheses
that could be explored with the method. The method itself rests on
assumptions, discussed in \sect~\ref{sec:assumptions}, that can either be
under test or relied upon, depending on the case.

Some approximations implicit in the method make its numerical results
semi-quantitative with respect to an analysis based on the full probability
calculus. The results should not be off by more than a few percentages for
the normalized total activity and for the frequency density. This
imprecision, however, is compensated by the extreme computational cheapness
of the method, far less costly than a full probabilistic analysis. The
method is therefore very convenient at least for essaying hypotheses. We
hope to present an exact quantitative comparison with the full
probabilistic approach in a future publication.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-06T17:01:42+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Answering this kind of questions always engages an element of uncertainty.
We cannot say \enquote{the answer is such and such}; at best we can assign
degrees of reasonable belief -- that is, probabilities -- to the possible
answers. The assessment of such distributions of belief involves
experimental data, such as recordings of neuronal activity from specific
brain areas, and pre-data knowledge and hypotheses about biological
conditions and mechanisms. When we translate such pre-data beliefs into an
initial probability distribution we often simplify them in more or less
realistic ways, to make the analysis mathematically tractable. This is why
such initial probability distributions are called \enquote{models}.

The best experimental measurements of instantaneous neuronal activity use
remarkable technologies, but can still only record a very small sample of
neurons -- hundreds at most -- compared to the numbers that constitute a
functionally distinguished brain region. Many probabilistic models focus on
such samples only: they neglect that the recorded neurons are a sample from
a larger population. Such isolating assumptions sometimes escape attention,
being subtly hidden in the mathematics. Some probabilistic models try to
take unrecorded neurons into account, but they do so by describing each
neuron individually, thus becoming very complex
\citep{huang2015,battistinetal2017}. It would be useful to explore models
that operate in between, addressing the activity of the larger brain area
whence the sample comes, but \emph{collectively}, that is, without asking
about individual neuronal details. Such intermediate models would be useful
at least for preliminary investigations, to help us to decide which
hypotheses to discard and which to consider for more complex and costly
models, or to suggest new hypotheses.

% \mynote{[Yasser:] In the present work we aim to address the issue of
%   building probability distributions over a large population using
%   recording from subpopulations by considering the distribution of total
%   population activity as an example. [} 
In the present work we propose such an intermediate probabilistic model. It
answers this question: How much was the \emph{total} activity of a large
neuronal population, given the observation of the activity of a very small
sample thereof? This model addresses a larger brain area, avoiding the
assumption of isolation of the sample; and by focusing on the total
activity, rather than the activity of individual neurons, it remains simple
and numerically tractable.

The model can be viewed as a straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b_r1996,grandy1980}
to build initial probability distributions has been used for different
kinds of investigations about the neuronal activity of various brain areas
and about other phenomena of importance to the neurosciences, for example
gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.
However, our model can also be viewed as an approximation of a more
detailed model within a fully Bayesian approach; we shall constantly keep
this point of view in mind in our discussion.

We illustrate possible uses of our model in \sect~\ref{sec:application}, by
applying it to two concrete data sets:
\begin{enumerate*}[label=(\alph*)]
\item the activity of 65 neurons recorded for 20 min from a rat's Medial
  Entorhinal Cortex \citep{stensolaetal2012},
\item the activity of 159 neurons recorded for 15 min from a macaque's
  Motor Cortex \citep{rostamietal2016_r2017}.
\end{enumerate*}
For each data set the model gives us the most plausible frequencies of all
levels of total activity of a much larger population, 1\,000--10\,000
neurons in size, during the recording. For example it can tell us that 60
out of 10\,000 neurons were most likely active during 1\% of the recording
time (though not necessarily always the same 60), 250 neurons out of
10\,000 were active during 0.4\% of the recording time, and so on. The
precise meaning of this frequency distribution is explained in
\sect~\ref{sec:model}. For the two example data sets, the guessed frequency
distributions for the larger population are distinctly different from those
for the sample. For the first data set the frequency distribution for the
larger population has two very distinct modes, both at low
activities %($0.7\%$ and $2.6\%$ of simultaneously active neurons
% within $3\;\mathrm{ms}$,
(see \fig~\ref{fig:plots_5mom}), whereas the frequency distribution for the
sample is monotonically decreasing with its maximum at zero activity. The
frequency distribution for the second data set presents one prominent
shoulder in its low-activity mode. These results show that the proposed
model can lead to the formulation or preliminary assessment of interesting
hypotheses, as we will illustrate with toy examples in
\sects~\ref{sec:application} and~\ref{sec:marginalization}. Note that it is
not possible to guess these features of the larger population by applying
the maximum-entropy method \emph{at the level of the sample alone}.

@@

Our model also solves a methodological problem in the use of
maximum-entropy methods to assess the \enquote{cooperativity},
\enquote{interaction}, or \enquote{synchrony} in neuronal activity, for
example studying its pairwise correlations and correlations of higher order
\citep[see for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
When the difference in size between a large population and a sample thereof
is too large, the sufficiency of correlations of some order for the sample
implies the \emph{lack} of sufficiency of correlations of the same order
for the larger population, and vice versa. Maximum-entropy applications at
the sample level can therefore deceive us in questions regarding the
cooperativity of the larger population. Our model addresses directly the
latter instead. We explain this point more precisely in
\sect~\ref{sec:marginalization}.

@@

\section{Quantifying the importance of higher-order correlations:\\
  the limitations of models at the sample level}
\label{sec:marginalization}

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been used as a way of
quantifying the \enquote{cooperativity} \citep[\eg][]{gersteinetal1985} or
\enquote{interaction}
\citep[\eg][]{martignonetal1995,schneidmanetal2006,shlensetal2006} or
\enquote{synchrony} \citep[\eg][-- we're only citing early papers using
these terms]{bohteetal2000,amarietal2003} of neuronal activity. Most, if
not all, such applications use the method \emph{at the sample level}. In
this section we discuss how our proposed application bears on this kind of
quantification and compare it with the traditional application at the
sample level. But such a comparison involves some methodological
caveats which we wish to discuss first.% For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our
% discussion holds for any sets of statistics.

@@

This fact is important for our analysis. If, say, means and pairwise
correlations seem informationally sufficient for a particular sample from a
brain area, then they may well not be sufficient for the larger population
of neurons constituting that area, and vice versa. So if we are interested
in the \enquote{cooperativity} or \enquote{interaction} of a brain area, it
is unreliable to use a maximum-entropy distribution constructed only for a
sample
thereof. % \citep[contrast this with some literature, \eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}
The model presented here avoids this problem because the maximum-entropy
method is applied to obtain the distribution of the larger population, not
of the sample alone.

@@

Compare the results above with those obtained by applying the model at the
sample level, that is, with $N=n$:
\begin{equation}
  \label{eq:diff_suff_n}
  \begin{split}
  \ydi_{n}(M_{4},M_{2}) &= 280\;\nat = 1\,220\;\hart,
\\    \ydi_{n}(M_{5},M_{4}) &= 3.3\;\nat = 1.4\;\hart.
\end{split}
\end{equation}
The data are 1\,220 orders of magnitude more probable conditional on four
moments than conditional on two, and $10^{1.4}=25$ times more probable
conditional on five moments than on four moments.

@@

The crucial point is our belief distribution for the activity of the
sample conditional on the activity of the larger population, given by
formula~\eqref{eq:conditional_hypergeometric} in
appendix~\ref{sec:derivation_maxent_fullnet}, reproduced here:
\begin{equation}
  \labelbis{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
    \binom{\yA}{\ya}\binom{N-\yA}{n-\ya}\binom{N}{n}^{-1}
\equiv \binom{n}{\ya}\binom{N-n}{\yA-\ya}\binom{N}{\yA}^{-1}
\end{equation}
leading to the equality of factorial
moments~\eqref{eq:equal_m_factorialmoments}, on which our model rests.
This conditional probability, characteristic of \enquote{drawing without
  replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968},
ensues when our degree of belief that the $i$th unit (ball or neuron) will
be drawn has the same value for all $i \in \set{1,\dotsc,N}$. Thus,
consider the pool of all neurons which we believe -- with equal degree for
each neuron -- could have been recorded. $N$ is the size of that pool. $N$
therefore depends on factors such as: the shape, dimensions, and technical
specifications of the recording probe; the inaccuracy in the insertion of
the probe, leading for example to slightly different insertion points or
angles; the density of neurons around the probe. But it also depends on the
homogeneity of the brain region where the recording was made: if we believe
that it doesn't matter whether the probe had been inserted in some other
point of the same brain region, then
formula~\eqref{eq:conditional_hypergeometric} is appropriate. $N$ can
therefore only be assessed case by case.

@@

A rigorous assessment would involve a more expensive, full Bayesian
calculation; but such calculation would make the maximum-entropy
method superfluous. We discuss this final point in the next section.

@@

The possible uses of the model here presented depend on the assumptions and
approximations which define it: three main assumptions and two main
approximation all in all. These become clear within a full-fledged
probabilistic approach. We summarize them here and give a more mathematical
sketch in appendix~\ref{sec:inference_prob}.

@@

The second assumption, typical of maximum-entropy in general, concerns the
form of the probability distribution for the long-run frequencies ensuing
from the first assumption. It is taken to be either an \enquote{entropic
  prior} \citep{skillingetal1984,rodriguez1991,neumann2007}, which takes
into account the multiplicity of long-run frequencies, or a prior for a
model with sufficient statistics, which expresses the sufficiency of the
moments under study \citep{portamana2017}. This assumption is crucial for
deriving the maximum-entropy based on with the Shannon entropy. A different
density leads to alternative maximum-entropy methods, for example based on
the Burg entropy \citep{burg1975} if a Dirichlet density is used
\citep{jaynes1986d_r1996}. The entropic prior depends on a reference
distribution and on a coefficient $L$ which determines the sharpness of the
prior's peak. The computations of the preceding sections were insensitive
to the reference distribution in their significant digits. We used three
biologically reasonable alternatives: a flat distribution, one linearly
decreasing with $\yA$, and a normal one with a broad peak at low
activities.

@@

Finally, the maximum-entropy method is typically equivalent to
approximating the number $T$ of time bins, the prior coefficient $L$, and
the ratio $T/L$ to infinity. These approximations lead to the
identification of the four distributions listed at the end of
\sect~\ref{sec:model}: the recorded frequency distribution becomes equal to
the long-run one, and also to the probability distribution for the activity
at recorded and new time bins. In our application it also leads to the
equality of the hypergeometric
distribution~\eqref{eq:conditional_hypergeometric} and the conditional
frequencies of the sample activity $\ya$ given $\yA$. The goodness of this
approximation decreases as the ratio between the number of possible joint
states and the number of bins -- roughly $nN/T$ -- increases, and also as
observed quantities approach their mathematical bounds, such as some
frequencies $\yff_{\ya}=0$. Without these approximations it is still
possible to interpret the maximum-entropy
distribution~\eqref{eq:app_maxent_pop_first} as the mode of the probability
distribution for the recorded frequencies $\yF$, that is, the most probable
recorded frequency.

@@

The $T= \infty$ approximation can break our model for specific
distributions of sample frequencies, if too many moments are used. For
example take $n=2$ with frequency distribution $\yf=(0.5,0,0.5)$ for the
activities $\ya=0,1,2$; that is, the sample is silent half of the time and
completely active half of the time. The first two normalized factorial
moments have both value $0.5$, and together they uniquely determine $\yf$.
If we now take $N=3$, it can be proved that
\eqn~\eqref{eq:app_maxent_pop_first}, constrained on those two moments, has
no solution. The reason is clear from a sampling perspective: if there is
zero probability of sampling one active neuron, then the larger population
cannot have any active neurons, and there must be zero probability of
sampling two active ones. But this reasoning holds only for each single
time bin, not for the averages over time bins. The $T=\infty$ approximation
identifies the two instead. Outside this approximation there are no
contradictions because the conditional frequencies and conditional
probabilities are kept distinct.

@@

The reason why we record neurons from a brain region is to have an idea of
the activity of the other neurons in that region, similarly to survey
sampling. If we thought that \emph{precisely} the recorded ones were
exceptions or completely unrelated to the others surrounding the recording
probe, our recording would serve little purpose. Probabilistic models that
allow us to make direct inferences about the larger population are
therefore useful. But they become complex and sensitive if we ask for very
detailed inferences.

We have presented a model that allows us to make direct inferences about a
larger population from very small samples. Its inferences regard only the
frequencies of the total activity of the larger population, but such
limited inferential scope makes the model computationally very cheap.

Yet, and most important, this model leads to inferences that are not at all
trivial and not easily discernible by looking at the sample. We showed this
by applying the model to two real data sets.

By \enquote{inference} we mean the quantification of a degree of belief
about possible observations, based on specific assumptions; not the
response of an oracle. The model here presented gives us a \emph{forecast}
based on specific assumptions discussed in \sect~\ref{sec:assumptions}, and
therefore has two main uses, just like any other probabilistic model. If
our assumptions are only hypotheses under testing, then the subsequent
confirmation or confutation of the forecast will increase or decrease our
confidence in those hypotheses. If we strongly trust our assumptions, on
the other hand, then we rely on the forecast and even use it in place of
actual observations, especially if the latter are difficult to make. (In
fact, we typically shift from the first to the second use.) In the
preceding sections we have given examples of additional hypotheses that
could be explored with the present model. The assumptions of the model
itself (\sect~\ref{sec:assumptions}) can either be under test or relied
upon, depending on the case.

Some approximations implicit in the model make its numerical results
semi-quantitative with respect to an exact analysis. The results should not
be off by more than a few percentages for the normalized total activity and
for the frequency density. This imprecision, however, is compensated by the
extreme computational cheapness of the model, far less costly than an exact
analysis. The model is therefore very convenient at least for essaying
hypotheses. We hope to present an exact quantitative comparison with the
full probabilistic approach in a future publication.
